Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-07-10 manuscript/manuscript.tex
Incorporating Jun Yan's comments
Reduced chapters for re-org

\section{Introduction} 
\label{sec:intro}
In the early 21st century,
the open data movement began to take shape, driven by the
fundamental belief that freely accessible data can transform 
both societies and governments. This movement champions the principles
of transparency, innovation, and public engagement. 
A landmark in this journey was the launch of the United States'
\href{https://www.data.gov}{Data.gov} portal in 2009, a pioneering
platform in making government data widely accessible. Shortly after,
the European Union followed suit, unveiling its
\href{https://data.europa.eu/euodp}{Open Data Portal} in 2012, further
cementing the movement's global reach. Furthermore, the World Bank's Open
Data initiative, initiated in 2010, stands out as a comprehensive
repository for global development data, available at
\href{https://data.worldbank.org}{World Bank Open Data}. 
These initiatives represent significant strides in democratizing data, 
in breaking barriers that once kept valuable information 
on government performance in silos. Their collective impact 
is profound, extending beyond mere data sharing to 
fostering a culture of openness that benefits individuals, 
communities, governments, and economies worldwide 
\citep{barns2016mine, wang2016adoption}.

@@

The paper is organized as follows:
Section~\ref{sec:data} offers a brief review of the history of the 311
system and the long-term trends presented via a 10-year analysis.
Section~\ref{sec:issues} offers a 
general discussion of data cleansing issues impacting data 
quality and curation efficiency. Section~\ref{sec:structural} examines
the technical dimensions of the dataset, the structural issues. 
In Section~\ref{sec:datatypes} we examine the data fields for compliance 
with the Data Dictionary. Section~\ref{sec:blanks} looks at the 
data fields as regards missing, blank, or N/A entries. Section~\ref{sec:domain} 
explores how data fields comply with a domain of legal 
or acceptable values. Section~\ref{sec:inconsistencies} deals 
with the important issues surrounding logical inconsistencies 
and concerning patterns in the data. Section~\ref{sec:precision} 
explores the issue of precision versus accuracy. Section~\ref{sec:duplicates}
identifies duplicate and redundant data fields. Section~\ref{sec:improvements} provides 
actionable suggestions for mitigating or resolving identified issues. 
Following this, Section~\ref{sec:protocol} outlines a series of general 
principles for the release of open data, drawing from our findings. The 
paper concludes with a discussion and recommendations in 
Section~\ref{sec:discussion}, encapsulating the key insights and 
implications of our research.

\section{NYC 311 Service Request Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results.

Despite its success, the 311 system faces several challenges:
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, excluding personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usagefor numerous
third-party users. These challenges are addressed by various agency
open data managers and the
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system.

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping city governance and community
engagement. This open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID-19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
improving responsiveness for city agencies, and addressing 
homelessness through initiatives like the Homeless
Outreach and Mobile Engagement Street Action Teams (HOME-STAT).

To begin, we looked at a 10-year period (2014-2023) to
observe the data over a longer time-frame; see
Figure~\ref{fig:10-yr-monthly}. During that time-frame,
the number of 311 SRs grew by nearly 50\%.

%\begin{figure}[tbp]
%	\centering
 % 	\includegraphics[width=0.8\textwidth]{10-year-trend-monthly.pdf}
 %	\caption{10-year (2014-2023) Monthly SR Counts}
 % 	\label{fig:10-yr-monthly}
%\end{figure}

%\begin{figure}[tbp]
%	\centering
%	\includegraphics[width=0.8\textwidth]{SRs_by_Agency.pdf}
 % 	\caption{SR counts by Agency with Cumulative Percentage}
%	\label{fig:SRcountbyAgency}
%\end{figure}

Figure~\ref{fig:SRcountbyAgency} displays SR counts by Agency with
cumulative percentage. Identifying the 
responsible agency and type of complaint is essential 
for troubleshooting discrepancies. 
This identification also assists in determining whether an issue is 
specific to an Agency or indicative of a broader, systemic problem. 
Six agencies account for 90\% of the complaints. These are the New York
Police Department (NYPD), Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).

\begin{comment}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=0.8\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint\_type(s) and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

\begin{table}[tbp]
  \centering
  \caption{Noise-related complaints\_type(s) by count with Agency}
  \label{tab:noisecomplaints}
  \begin{tabular}{@{}lS[table-format=7.0,round-mode=places,
    round-precision=0]S[table-format=2.2,round-mode=places,
    round-precision=2]l@{}} % 'l' for left-aligned, 'S' for siunitx number alignment
    \toprule
    \textbf{complaint\_type} & \textbf{Count} & \textbf{Percentage} & \textbf{Agency} \\ 
    \midrule
    Noise - Residential        & 675502 & 10.56 & NYPD  \\ 
    Noise - Street/Sidewalk    & 300507 &  4.70 & NYPD  \\ 
    Noise - Commercial         & 128892 &  2.02 & NYPD  \\ 
    Noise - Vehicle            & 115956 &  1.81 & NYPD  \\ 
    Noise                      & 100413 &  1.57 & DEP   \\ 
    Noise - Helicopter         &  85345 &  1.33 & EDC   \\ 
    Noise - Park               &  16620 &  0.26 & NYPD  \\ 
    Noise - House of Worship &   2757 &  0.04 & NYPD  \\ 
    \bottomrule
  \end{tabular}
\end{table}
\end{comment}

Figure~\ref{fig:SR_complaints} shows that the top 20 complaint 
types account for 70\% of the total 220 different types. Notably, 
these top complaints include several variations of noise-related 
issues, which collectively represent 22\% of all complaints, which
when consolidated is the most frequently occurring complaint type.

\section{Data Cleansing Issues} 
\label{sec:issues}
What is data cleansing?  Wikipedia 
\href{https://en.wikipedia.org/wiki/Data_cleansing}{Data Cleansing} offers a good definition.

@@

Many quality criteria are required in order to process high-quality 
data. These include:

\begin{itemize}
	\item Data validation - this effort can span a number of criteria
	\begin{itemize}
		\item Mandatory fields: Certain data fields cannot be empty.
		\item Data-types: Certain fields must be of the correct type, e.g. numeric, 
		character, date, 5 numeric digits, etc. Typically these are specified 
		in a Data Dictionary.
		\item Domain compliance:  Many data fields must adhere to a specific 
		domain of values, e.g. statuses, state names, zip codes, gender. 
	\end{itemize}   
	\item Structural errors to include naming conventions, 
	 fields not in the Data Dictionary, or inconsistent data entry. For example 
	 intermixing blanks, spaces, NA, N/A, and \textless{}NA\textgreater{} 
	 all to indicate the absence of data.
	\item Redundant, unnecessary, or irrelevant fields
	\item Logical inconsistencies such as related fields that violate the 
	nature of that relationship, e.g. a ``due date'' that is before the ``created date''
	\item Accuracy versus precision
\end{itemize}

This analysis effort is to identify the presence of such errors; not to correct them. 
That effort would be undertaken only after an investigation as to the why
and how such errors came about, and a discussion as to whether or not it 
is even an ``error''.

\section{Structural Issues}
\label{sec:structural}
Structural issues refers to how data is organized, formatted, 
or structured within a dataset. Structural issues can make 
it difficult to analyze the data effectively. Here are some 
characteristics of the 311 SR data set:

\begin{itemize}
	\item There are 47 columns of data for each row, exportable as a CSV file.
	\item There are four date fields (created, closed, updated, due).
	\item There are three borough fields; two of which appear to be duplicates.
	\item Two zip code fields, but not duplicates
	\item Seven street fields; one pair of which appear to be duplicates
	\item Two Police Precinct fields; not duplicates
	\item In addition to incident\_address, there are five additional location fields: 
	lat/long, street\_name, landmark, NY State plane, and Block \#
	\item One free-form text field, resolution\_description, which 
	supports 934 characters of input, including commas and special characters
\end{itemize}

\textbf{Issue: Fields not in the Data Dictionary} The 311 SR 
\href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
 identifies 41 data columns (fields) along with related information 
 for each column. However, when downloaded the data contains 
 six fields are not included in the Data Dictionary. However, these
 fields do show up on the portal Column Manager widget
 as ``@computed\_region\_xxxx\_xxxx''. To some extent one can infer 
 what the these @computed\_fields are and how they are determined but
 not with certainty. Most importantly, can these fields be reliably 
 used for analytical purposes?  These six fields are:

\begin{itemize}
	\item zip\_codes
	\item community\_districts
	\item borough\_boundaries
	\item city\_council\_districts
	\item police\_precincts
	\item police\_precinct 
\end{itemize}

\section{Validating data types}
\label{sec:datatypes}
Fortunately, both the Data Dictionary and the portal column manager 
clearly indicate the epected data type. The following fields were 
checked for their specified data types:

\begin{itemize}
	\item created\_date, closed\_date, due\_date, and resolution\_action\_updated\_date: 
	All are all in proper date format.

	\item zip\_codes and incident\_zip: All are 5 numeric digits except 
	for two non-numeric entries for incident\_zip (``na'' and ``N/A'').

	\item x\_coordinate\_state\_plane \&  y\_coordinate\_state\_plane 
	(a NY State geo-location system): All are all numeric.

	\item latitude \& longitude: Both are numeric.

	\item community\_districts, borough\_boundaries, 
	city\_council\_district, police\_precinct, and police\_precincts: All are numeric.

	\item Free-form text fields that contain a ``,'' (e.g. incident\_address 
	\& resolution\_description) are properly enclosed in quotation marks.
\end{itemize}

\section{Blank \& N/A data}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to see if the SRs were
closed before or after their due\_date, you would be challenged 
as 99.57\% of the due\_date field is blank. When counting 
the various fields for blank or N/A values, the fields appear 
to divide into three groups: \textbf{Mostly Empty}, \textbf{Partially Empty}, 
and \textbf{Few/None Empty}. The Mostly Empty category 
ranges from 93-99.9\% blank and includes such fields as 
taxi\_company due\_date, pickup\_location, and 
landmark. The Partially Empty includes such fields as location\_type, borough, 
and cross\_street. And the Few/None Empty includes created\_date, 
complaint\_type, agency, and status. In some cases it may make 
sense to inquiry as to why some fields are frequently blank. Here is a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Some, and Few categories.

%\begin{figure}[tbp]
%	\centering
 % 	\includegraphics[width=\textwidth]{BlankFields.pdf}
%	\caption{Number and Percentage of Empty/Blank Entries}
%	\label{fig:blank_fields}
%\end{figure}

 \section{Validating Data for Acceptable Values}
 \label{sec:domain}
Any analytic effort needs to be aware of fields which 
contain invalid values. Typically such fields would need to be
identified and removed from the analysis. Here are some results
of field validation.

\begin{itemize}
	\item Latitude and Longitude fields fell within the geographic 
	boundaries of New York City.

	\item The unique\_key field was in fact unique.

	\item Unfortunately, very few domains of acceptable 
	values are specified in the Data Dictionary. The below 
	fields were tested as complying with their domain as 
	determined by public usage and historical datasets:
	\begin{itemize}
		\item address\_type
		\item status
		\item borough, borough\_boundaries, \& park\_borough 
		\item data\_channel
		\item vehicle\_type
		\item city\_council\_district
	\end{itemize}	
\end{itemize}

%\subsection{Issues with Zip Codes}
\label{sec:zipcodesissues}
 \textbf{Issues with Zip Codes} Some fields proved to be problematic 
 as to complying with a domain of legal values. For example, all zip codes 
 (two fields: zip\_codes and incident\_zip) should be valid as defined 
 by the USPS database which contains 37,946 valid zip codes. However,
 the computed field zip\_codes proved problematic with 
58\% (3.6 million) of the field entries being invalid. And the 
incident\_zip field, while having only .07\% invalid entries, still contains
4163 such errors. The breakdown of the invalid entries in the zip\_code, sorted by 
Agency shows that the distribution by percentage mirrors the overall 
breakdown of SRs by Agency, potentially indicating a systemic problem.

%\subsubsection{Case Study: Noise Complaints by Zip Code}
\label{sec:case-study-zip-codes}
\textbf{Case Study: Noise Complaints by Zip Code} Scenario: The NYC 
Office of Nightlife (ONL) wants to know ``What are the top 10 
zip codes for Noise Complaints over the last two years?'' ONL wants
to assess the impact of the recent NYC effort promoting a vibrant nightlife 
in NYC while seeking to ease strained relations between bar 
and club owners. Let's assume that this analysis 
uses the zip\_codes field, one of the (six) computed fields known to 
have validity problems. We then repeat the analysis with the more 
accurate incident\_zip field. We subject these two analysis effort 
to validation against the USPS zip code database.

\begin{comment}	 
\begin{table}[tbp]
    \centering
    \caption{Comparison of Top Ten Zip Codes Lists}
	    \begin{tabular}{@{}lS[table-format=6.0,round-mode=places,
	    round-precision=0]c@{\hskip 0.5cm}@{}lS[table-format=6.0,
	    round-mode=places,round-precision=0]c@{}}
		\toprule
	 	\multicolumn{3}{c}{\textbf{zip\_codes}} & \multicolumn{3}{c}{\textbf{incident\_zip}} \\
	      \cmidrule(r){1-3} \cmidrule(l){4-6}
	      \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} 
	      & \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} \\
	      \midrule
	        11275 & 104556 & FALSE & 10466 & 104562 & TRUE \\
	        12420 & 27503 & TRUE & 10023 & 27972 & TRUE \\
	        12428 & 26564 & TRUE & 10031 & 25548 & TRUE \\
	        10935 & 25508 & FALSE & 10457 & 25066 & TRUE \\
	        10934 & 23448 & FALSE & 10453 & 24752 & TRUE \\
	        10931 & 22381 & TRUE & 10456 & 24751 & TRUE \\
	        10930 & 22121 & TRUE & 10452 & 22527 & TRUE \\
	        17613 & 21963 & FALSE & 10025 & 21705 & TRUE \\
	        10936 & 21707 & FALSE & 10458 & 21689 & TRUE \\
	        11606 & 21435 & FALSE & 10032 & 20622 & TRUE \\
	      \bottomrule
	    	\end{tabular}
 	\label{tab:zipcodes}
\end{table}
\end{comment}

As indicated, six out of ten zip\_codes are invalid, which corresponds closely 
with what is observed in the overall dataset (58\% invalid). Whereas 
the incident\_zip field is completely valid, in-line with the 
overall incident\_zip validation percentage of 99.04\%.

%\subsection{Issues with Police Precincts} 	
\label{sec:police-precincts}
A curious case exists when examining two (nearly) identical 
fields: police\_precincts and police\_precinct. Both of those fields 
are among the ``computed'' fields previously mentioned. Using 
\href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}
{NYPD Precinct listings} it is possible to determine what are valid 
police precincts. What we find is that both police\_precinct 
and police\_precincts  have  35\% invalid entries. Unfortunately, 
it's not the same invalid entries. For the police\_precincts field, 
there are 2,171,864 invalid entries (35\%). Similarly for the 
police\_precinct field, there are 2,171,778 invalid entries 
(also 35\%), but not the same invalid entries.

%\subsection{Issues with Community Boards}
\label{sec:communityboards}
Community Boards are the most local, grassroots form of 
City government; a vital connection between communities and 
elected officials and City agencies. Community Boards are often
used as a measure of equity of City services throughout the five 
boroughs. In the 2022-2023 dataset there are 27,276 invalid community\_board 
entries which represents 0.43\% of non-blank data. The distribution 
of invalid Community Boards by Agency is not consistent with 
the overall SR Agency distribution. This distribution indicates that 
there are likely issues at key Agencies, in this case Taxi \& Limo 
Commission (TLC), Parks \& Recreation, etc.

%\subsection{Issues with Community Districts}
\label{sec:communitydistrict}
Another of the ``computed'' fields is community\_districts. Community Districts 
are the boundaries for the Community Boards, but differ as they
are used by the Department of City Planning (DCP) for purposes 
of environmental, socio-economic, and demographic 
purposes. It is a geographical division rather than a local 
government division. Due to how the community\_district data 
is formatted, it is not possible to establish validity. However, 
it is possible to determine that the dataset contains 72 unique 
entries, while there are only 59 valid Community Districts.

%\subsection{created\_date and closed\_date(s) -- Negative Duration}
\label{sec:negativeduration}
For the created\_date and closed\_date fields, you might 
expect that these two fields are automatically populated by the 
SR application software; associated with setting an SR status 
to ``new'' or ``closed''. Unfortunately, that does not appear 
to be the case as there are several anomalies with the various date 
fields, including:

@@

%\section{Protocol Suggestions} \label{sec:protocol}

>>>>

\section{INTRODUCTION} 
\label{sec:intro}
In the early 21st century, the open data movement began 
to take shape, driven by the fundamental belief that 
freely accessible data can transform both societies and 
governments. This movement champions the principles
of transparency, innovation, and public engagement. 
A landmark in this journey was the launch of the United States'
\href{https://www.data.gov}{Data.gov} portal in 2009, a pioneering
platform in making government data widely accessible. Shortly after,
the European Union followed suit, unveiling its
\href{https://data.europa.eu/euodp}{Open Data Portal} in 2012, further
cementing the movement's global reach. Furthermore, the World Bank's Open
Data initiative, initiated in 2010, stands out as a comprehensive
repository for global development data, available at
\href{https://data.worldbank.org}{World Bank Open Data}. 
These initiatives represent significant strides in democratizing data, 
in breaking barriers that once kept valuable information 
on government performance in silos. Their collective impact 
is profound, extending beyond mere data sharing to 
fostering a culture of openness that benefits individuals, 
communities, governments, and economies worldwide 
\citep{barns2016mine, wang2016adoption}.

@@

\jy{How about putting all these sections into one section on case
  study? They can be subsections in the merged section.
  After that, we can have a section on recommendations for open
  data curation based on what we learn from the case study. For each
  recommended principle, we can link to the specifics of the case
  study.}

The paper is organized as follows:
Section~\ref{sec:data} `\textit{NYC 31 Service Request Data} offers a 
look at the history of the 311system. Long-term trends are 
presented in a 10-year data analysis (2014-2023).
Section~\ref{sec:issues} \textit{Data Cleansing Issues} offers a 
discussion of specific data cleansing issues impacting data quality and 
curation efficiency. These issues include structural issues, compliance 
with the Data Dictionary, and missing, blank, or N/A entries. We also
explore field compliance with a domain of legal or acceptable values.
Next we deal with several important issues surrounding logical inconsistencies 
and concerning patterns found in the data. We also explore an issue of precision 
versus accuracy. This section concludes by identifying duplicate/redundant data fields
and Data Dictionary observation. Section~\ref{sec:recommendations} 
\textit{Recommendations} provides actionable suggestions for 
mitigating or resolving the identified issues. Section~\ref{sec:discussion} ``Discussions'' 
concludes with an encapsulation of key insights and implications of our research.

\section{NYC 311 Service Request Data} 
\label{sec:data}
The NYC 311 service, a critical component of New York City's public
engagement and service response framework, serves as a centralized hub
for non-emergency inquiries and requests. Introduced in 2003, the NYC
311 system was designed to streamline the city's response to
non-emergency issues, ranging from noise complaints to street
maintenance requests. Initially a phone-based call center, the system
evolved into a comprehensive data management platform handling
millions of requests annually. Key milestones since its launch in 2003
include the addition of online and mobile app channels in 2009, a
record high of 348,463 monthly service requests in August 2020 due to
the COVID-19 pandemic, the 2021 expansion to include the MTA's city
subway system, and a record 3.23 million service requests in
2023. Today, the NYC 311 data system manages over 3 million service
requests per year. This data is publicly accessible through the NYC Open Data
Portal, which provides tools for querying, grouping, aggregating,
geo-mapping, visualizing, and exporting results. The
\href{https://www.nyc.gov/content/oti/pages/}{NYC Office of Technology
  and Innovation (OTI)}, which provides technological support for the
open data system.

\begin{comment}
Despite its success, the 311 system faces several challenges:
data timeliness, accuracy, and consistency, difficulties in
correlating data over long periods, excluding personally
identifiable information (PII), integration with stand-alone systems
at selected NYC agencies, and managing API usagefor numerous
third-party users. These challenges are addressed by various agency
open data managers and 
\end{comment}

The impact of NYC 311 data extends beyond operational efficiency; it
has become instrumental in shaping city governance and community
engagement. This open data not only ensures governmental transparency
but also empowers researchers, civic developers, and the general
public. The data has been pivotal in providing advice on shelters
during emergencies, handling inquiries during the COVID-19 pandemic,
enforcing standards between landlords and tenants, reallocating taxi
routes based on analyses by the Taxi and Limousine Commission (TLC),
and improving responsiveness across City Agencies

To begin, we looked at a 10-year period (2014-2023) to
observe a longer time-frame; see
Figure~\ref{fig:10-yr-monthly}. During that time-frame,
the number of 311 SRs grew by nearly 50\%.

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=0.8\textwidth]{10-year-trend_yearly.pdf}
 	\caption{10-year (2014-2023) Monthly SR Counts}
  	\label{fig:10-yr-monthly}
\end{figure}

Identifying the responsible agency is essential 
for troubleshooting discrepancies. This identification also assists in 
determining whether an issue is specific to an Agency or indicative 
of a broader, systemic problem. Six agencies account for 90\% of 
the complaints: New York Police Department (NYPD), 
Housing Preservation and Development (HPD),
New York City Department of Sanitation (DSNY), Department of
Transportation (DOT), Department of Environmental Protection (DEP),
and Department of Parks and Recreation (DPR).

\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.8\textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

\begin{comment}
\begin{table}[tbp]
  \centering
  \caption{Noise-related complaints\_type(s) by count with Agency}
  \label{tab:noisecomplaints}
  \begin{tabular}{@{}lS[table-format=7.0,round-mode=places,
    round-precision=0]S[table-format=2.2,round-mode=places,
    round-precision=2]l@{}} % 'l' for left-aligned, 'S' for siunitx number alignment
    \toprule
    \textbf{complaint\_type} & \textbf{Count} & \textbf{Percentage} & \textbf{Agency} \\ 
    \midrule
    Noise - Residential        & 675502 & 10.56 & NYPD  \\ 
    Noise - Street/Sidewalk    & 300507 &  4.70 & NYPD  \\ 
    Noise - Commercial         & 128892 &  2.02 & NYPD  \\ 
    Noise - Vehicle            & 115956 &  1.81 & NYPD  \\ 
    Noise                      & 100413 &  1.57 & DEP   \\ 
    Noise - Helicopter         &  85345 &  1.33 & EDC   \\ 
    Noise - Park               &  16620 &  0.26 & NYPD  \\ 
    Noise - House of Worship &   2757 &  0.04 & NYPD  \\ 
    \bottomrule
  \end{tabular}
\end{table}
\end{comment}

Figure~\ref{fig:SR_complaints} shows that the top 20 complaint 
types account for 70\% of the total 220 different types. Notably, 
these top complaints include several variations of noise-related 
issues, which collectively represent 22\% of all complaints, the 
most frequently occurring type of complaint.

\begin{figure}[tbp]
 \centering
  \includegraphics[width=0.8\textwidth]{SR_by_Complaint_Type.pdf} 
  \caption{Top 20 complaint\_type(s) and Cumulative Percentage} 
  \label{fig:SR_complaints}
\end{figure}

\section{Data Cleansing Issues} 
\label{sec:issues}
What is data cleansing?  Wikipedia 
\href{https://en.wikipedia.org/wiki/Data_cleansing}{Data Cleansing} 
offers this definition.

@@

Many quality criteria are utilized in order to process high-quality 
data, including:
\begin{comment}
\begin{itemize}
	\item Data validation - this effort can span a number of criteria

	\begin{itemize}
		\item Mandatory fields: Certain data fields cannot be empty.

		\item Data-types: Certain fields must be of the correct type, e.g. numeric, 
		character, date, 5 numeric digits, etc. Typically these are specified 
		in a Data Dictionary.

		\item Domain compliance:  Many data fields must adhere to a specific 
		domain of values, e.g. statuses, state names, zip codes, gender.

	\end{itemize}   
	\item Structural errors to include naming conventions, 
	 fields not in the Data Dictionary, or inconsistent data entry. For example 
	 intermixing blanks, spaces, NA, N/A, and \textless{}NA\textgreater{} 
	 all to indicate the absence of data.

	\item Redundant, unnecessary, or irrelevant fields

	\item Logical inconsistencies such as related fields that violate the 
	nature of that relationship, e.g. a ``due date'' that is before the ``created date''

	\item Accuracy versus precision
\end{itemize}
\end{comment}

This purpose of this analysis effort is to identify the presence of 
errors; not necessarily to correct them. A correction effort should
be undertaken only after an investigation as to the why and how 
such errors came about, and a discussion as to whether or not it 
even is an \textit{error}.

\subsection{Structural Issues}
\label{sec:structural}
Structural issues refers to how data is organized, formatted, 
or presented within a dataset. Structural issues can make 
it difficult to analyze the data effectively. Here are some 
characteristics of the 311 SR data set:

\begin{itemize}
	\item There are 47 columns of data for each row, exportable as a CSV file.

	\item There are four date fields (created, closed, updated, due).

	\item There are three borough fields; two of which appear to be duplicates.

	\item Two zip code fields, but not duplicates

	\item Seven street fields; two pair of which appear to be duplicates

	\item Two Police Precinct fields; not duplicates

	\item In addition to the incident\_address, there are five additional location fields: 
	lat/long, street\_name, landmark, NY State plane, and Block \#

	\item One free-form text field, resolution\_description, which 
	supports 934 characters of input, including commas and special characters
\end{itemize}

The 311 SR \href{https://data.cityofnewyork.us/api/views/erm2-nwe9/files/b372b884-f86a-453b-ba16-1fe06ce9d212?download=true&filename=311_ServiceRequest_2010-Present_DataDictionary_Updated_2023.xlsx}{Data Dictionary}
 identifies 41 data columns (fields) and related information 
 for each column. However, when downloaded the data reveals 
 six additional fields that are not included in the Data Dictionary. 
(Note that these fields do show up on the portal Column Manager widget
 as ``@computed\_region\_xxxx\_xxxx''.) To some extent one can infer 
 what the these @computed\_fields are and how they are determined, but
 not with certainty. Most importantly is the question, can these fields be reliably 
 used for analytical purposes?  These six ``computed'' fields are:

\begin{itemize}
	\item zip\_codes

	\item community\_districts

	\item borough\_boundaries

	\item city\_council\_districts

	\item police\_precincts

	\item police\_precinct 
\end{itemize}

\subsection{Blank \& N/A data}
\label{sec:blanks}
Understanding the absence of data by field is important 
when undertaking analysis. For example, if you wanted to 
determine if SRs were closed before or after their due\_date, 
you would be challenged as 99.6\% of the due\_date field is 
blank. When counting fields for blank or N/A values, they appear 
to divide into three groups: \textbf{Mostly Empty}, \textbf{Partially Empty}, 
and \textbf{Few/None Empty}.

The Mostly Empty category ranges from 93-99.9\% blank 
and includes such fields as taxi\_company due\_date, pickup\_location, and 
landmark. The Partially Empty includes such fields as location\_type, borough, 
and cross\_street. And the Few/None Empty includes created\_date, 
complaint\_type, agency, and status. In some cases it may make 
sense to inquiry as to why some fields are frequently blank. Here is a graphic 
depiction of total empty (blank \& N/As) for each fields illustrating
the grouping into the Most, Partial, and Few categories.

\begin{figure}[tbp]
	\centering
  	\includegraphics[width=\textwidth]{BlankFields.pdf}
	\caption{Number and Percentage of Empty/Blank Entries}
	\label{fig:blank_fields}
\end{figure}

\begin{comment}
 \subsection{Validating Data for Acceptable Values}
 \label{sec:domain}
Any analytic effort needs to be aware of fields which 
contain invalid values. Typically such fields would be
identified and removed from the analysis. Here are the results
of selected field validation.

\begin{itemize}
	\item Latitude and Longitude fields fall within the geographic 
	boundaries of New York City.

	\item The unique\_key field was in-fact unique.

	\item Unfortunately, very few domains of acceptable 
	values are specified in the Data Dictionary. The below 
	fields were tested as complying with their domain as 
	determined by public usage and historical datasets:

	\begin{itemize}
		\item address\_type
		\item status
		\item borough, borough\_boundaries, \& park\_borough 
		\item data\_channel
		\item vehicle\_type
		\item city\_council\_district
	\end{itemize}

\end{itemize}
\end{comment}

\label{sec:zipcodesissues}
 \textbf{Issues with Zip Codes:} Some fields proved to be problematic 
 as to their compliance with a domain of legal values. For example, all zip codes 
 (two fields: zip\_codes and incident\_zip) should be valid as defined 
 by the USPS database containing all 37,946 valid zip codes. However,
 the computed field zip\_codes proved problematic with over 
58\% (3.6 million) of the field entries being invalid. And while the more accurate 
incident\_zip field has only .07\% invalid entries, it still contains
4163 such errors; not insignificant. The breakdown of the invalid 
entries in the zip\_code, sorted by Agency shows that the 
distribution by percentage mirrors the overall breakdown of 
SRs by Agency, potentially indicating a systemic problem.

\label{sec:case-study-zip-codes}
\textbf{Case Study: Noise Complaints by Zip Code.}  Scenario: The NYC 
Office of Nightlife (ONL) wants to know ``What are the top 10 
zip codes for Noise Complaints over the last two years?'' ONL wants
to assess the impact of recent efforts promoting a vibrant nightlife 
in NYC while seeking to ease strained relations between bar 
and club owners. Let's assume that this analysis 
uses the zip\_codes field, one of the computed fields known to 
have validity problems. We then repeat the analysis with the more 
accurate incident\_zip field. Finally, we subject these two analysis effort 
to validation against the USPS zip code database.

\begin{table}[tbp]
    \centering
    \caption{Comparison of Top Ten Zip Codes Lists}
	    \begin{tabular}{@{}lS[table-format=6.0,round-mode=places,
	    round-precision=0]c@{\hskip 0.5cm}@{}lS[table-format=6.0,
	    round-mode=places,round-precision=0]c@{}}
		\toprule
	 	\multicolumn{3}{c}{\textbf{zip\_codes}} & \multicolumn{3}{c}{\textbf{incident\_zip}} \\
	      \cmidrule(r){1-3} \cmidrule(l){4-6}
	      \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} 
	      & \textbf{Zip Code} & \textbf{Count} & \textbf{Valid?} \\
	      \midrule
	        11275 & 104556 & FALSE & 10466 & 104562 & TRUE \\
	        12420 & 27503 & TRUE & 10023 & 27972 & TRUE \\
	        12428 & 26564 & TRUE & 10031 & 25548 & TRUE \\
	        10935 & 25508 & FALSE & 10457 & 25066 & TRUE \\
	        10934 & 23448 & FALSE & 10453 & 24752 & TRUE \\
	        10931 & 22381 & TRUE & 10456 & 24751 & TRUE \\
	        10930 & 22121 & TRUE & 10452 & 22527 & TRUE \\
	        17613 & 21963 & FALSE & 10025 & 21705 & TRUE \\
	        10936 & 21707 & FALSE & 10458 & 21689 & TRUE \\
	        11606 & 21435 & FALSE & 10032 & 20622 & TRUE \\
	      \bottomrule
	    	\end{tabular}
 	\label{tab:zipcodes}
\end{table}

As indicated, six out of ten zip\_codes are invalid, which corresponds closely 
with what is observed in the overall dataset (58\% invalid). Whereas 
the incident\_zip field is completely valid, in-line with the 
overall incident\_zip validation percentage of 99\%.

\label{sec:police-precincts}
\textbf{Issues with Police Precincts:} A curious case exists when 
examining two (nearly) identical fields: police\_precincts and 
police\_precinct. Both of those fields are among the computed fields 
previously mentioned. Using 
\href{https://www.nyc.gov/site/nypd/bureaus/patrol/precincts-landing.page}
{NYPD Precinct listings} it is possible to determine the valid 
police precincts. What we find is that both police\_precinct 
and police\_precincts  have about 35\% invalid entries. Unfortunately, 
it's not the same invalid entries. The police\_precincts field, 
has 2,171,864 invalid entries (35\%). Similarly the 
police\_precinct field has 2,171,778 invalid entries 
(also 35\%), but not the same invalid entries.

\label{sec:communityboards}
\textbf{Issues with Community Boards:} Community Boards are the most 
local, grassroots form of City government; a vital connection 
between communities, elected officials, and City agencies. Community 
Boards often service as a measure of the equity of City services 
throughout the five boroughs. In the 2022-2023 dataset there 
are 27,276 invalid community\_board entries representing 0.43\% 
of non-blank data. The distribution of invalid Community Boards 
by Agency is not consistent with the overall SR Agency distribution,
indicating that there are likely issues at key Agencies, in 
this case Taxi \& Limo Commission (TLC), DPR, etc.

\label{sec:communitydistrict}
\textbf{Issues with Community Districts:} Another of the computed fields 
is community\_districts. Community Districts 
are the geographic boundaries for the Community Boards, but differ as they
are used by the Department of City Planning (DCP) for environmental, 
socio-economic, and demographic purposes. It is a geographical 
division rather than a City government division. Due to how the 
community\_district data is formatted, it is not possible to 
establish validity. However, we can determine that the dataset 
contains 72 unique entries, while there are only 59 
valid Community Districts.

\label{sec:negativeduration}
\textbf{Issues with created\_date and closed\_date - Negative Durations:} For 
the created\_date and closed\_date fields, one might 
expect that these two fields are automatically populated by the 
SR application software; associated with perhaps setting an SR status 
to ``new'' or ``closed''. Unfortunately, that does not appear 
to be the case as there are several anomalies with the various date 
fields:

@@
